---
title: "Bellabeat Case Study"
author: "Coco Choo"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require("knitr")
```

Bellabeat, founded in 2013 by Urška Sršen and Sando Mur, is a tech-driven wellness company specializing in health-focused smart products designed to empower women. Tracking activity, sleep, stress, and reproductive health has enabled women to make informed decisions about their well-being.

By 2016, Bellabeat had expanded globally, launching multiple products available through online retailers and their e-commerce platform. The company combines traditional advertising with a strong focus on digital marketing, leveraging platforms like Google Search and various social media to engage consumers year-round.

To explore growth opportunities, Bellabeat’s marketing analytics team has been tasked with analyzing smart device usage data. This analysis aims to uncover user trends and the insights I discorver will provide strategic recommendations to enhance Bellabeat’s marketing efforts.

This study follows the six phases of the data life cycle which are:

1. Ask

2. Prepare

3. Process

4. Analyze

5. Share

6. Act


# Ask

Sršen has tasked you with analyzing smart device usage data to uncover trends in how consumers use non-Bellabeat smart devices. These insights will be applied to a selected Bellabeat product to inform strategic recommendations.

The primary stakeholders are:

* UrškaSršen: Bellabeat’s Co-founder and Chief Creative Officer

* SandoMur: Mathematician and Bellabeat’s Co-founder

* Bellabeat marketing analytics team: A team of data analysts 

# Prepare

The dataset used in this study can be found below:  

[FitBit Fitness Tracker Data](https://www.kaggle.com/datasets/arashnic/fitbit)

The datasets are accessible through Kaggle, licensed under the CCO Public Domain. The datasets contain personal fitness tracker from thirty fitbit users; for instance, daily activity, steps, minute-level output for physical activity, heart rate, and sleep monitoring. 

The datasets are in CSV format and are stored in long format with unique ID having multiple rows of data. For this study, I focused on activity dataset to discover trends in smart device usage.

# Process

Below are the datasets I will be using for this analysis:

* dailyactivity_merged (4.12.16-5.12.16)

I performed data cleaning using Excel and R. The following steps were taken for each dataset:

1. I sorted and filtered the data by ActivityDate

2. I checked for NULLs

3. I transformed the data type of ActivityDate column to Date type instead of Character type

```{r}
library(tidyverse)
library(dplyr)
library(tidyr)
library(ggplot2)
library(reshape2)

# Import data
setwd("C:\\Coco\\TTU\\Google Data Analytics Certification\\Bellabeat\\Fitabit_data_AM\\")

# Dataset I will be using for this analysis
activity <- read.csv("dailyActivity_merged.csv")
```

```{r}
# Check for null values
sum(is.na(activity))
```

```{r}
# Check the data types
str(activity)
```

```{r}
# Transform the ActivityDate column into Date type
activity$ActivityDate <- as.Date(activity$ActivityDate, format="%m/%d/%Y")
```

```{r}
# Preview the datasets
head(activity)
colnames(activity)
```

```{r}
# Unique participants in each dataset
n_distinct(activity$Id)
```

* Based on the result above, I observe that there are 33 unique participants in physical activity.

```{r}
# Check number of observations in each dataset
nrow(activity)
```

* After conducting data cleaning and formatting on the dataset, I printed out some summary statistics with information I will focus on in this study for descriptive analysis purposes.

```{r}
# Summary statistics
activity %>%
select(Calories,
       TotalSteps,
TotalDistance,
SedentaryMinutes) %>%
summary()
```

```{r}
# Summary statistics
activity %>%
select(VeryActiveMinutes,
FairlyActiveMinutes,
LightlyActiveMinutes) %>%
summary()
```
# Analyze and Share

Although summary statistics provide valuable insights about the data, it does not provide a clear picture of how each variable is related at one sight. Thus, I created some plots to visualize the findings, connecting the dots.

```{r}
# Relationship between steps and sedentary minutes
ggplot(data=activity, aes(x=TotalSteps, y=SedentaryMinutes)) + geom_point()
```

Based on the scatterplot displaying the relationship between Total Steps and Sedentary Minutes, I can observe some outliers with total steps above 25,000 but the overall, there appears to be a weak negative association between Total Steps and Sedentary Minutes. As Total Steps increases, Sedentary Minutes tend to decrease slightly, though the relationship is not very strong. One thing I notice is that the data points are densely clustered at lower Total Steps, around 0 to 10,000 and higher Sedentary Minutes, around 500 to 15,000, suggesting that most of the participants are relatively sedentary and do not have high step counts.

Then, I created a correlation matrix on numerical columns in activity dataset to investigate the correlation among the columns.

```{r}
# Focus on total steps, minutes and calories burnt
activity_num <- activity[c("Id", "ActivityDate", "TotalSteps", "VeryActiveMinutes", "FairlyActiveMinutes", "LightlyActiveMinutes", "SedentaryMinutes", "Calories")]

# Create a correlation matrix of the numerical data
corr <- round(cor(activity_num[sapply(activity_num, is.numeric)]), 2)
head(corr)

# Melt the correlation matrix
melt_corr <- melt(corr)
```

```{r}
# Visualize correlation matrix on a heatmap
ggplot(data = melt_corr, aes(x=Var1, y=Var2, fill=value)) + geom_tile(color = "white")
```

Now that I looked into relationship between several variables, I then proceeded to analyze further with Very Active Minutes and Calories to segment participants into three categories: Highly active, moderately active, and sedentary. 

The segmentation will help provide personalized Bellabeat marketing strategy based on the trends in the smart device usage. The process involves comparing the activity metrics I have selected to the mean values and then visualize them using a bar plot and a pie chart to get a clearer insight.

```{r}
# Find the average of very active minutes and calories
mean_very_active_minutes <- mean(activity$VeryActiveMinutes)
mean_calories <- mean(activity$Calories)
```

```{r}
# Segment users based on selected metrics relative to mean
activity$activity_level <- 
  ifelse(
      activity$VeryActiveMinutes >= mean_very_active_minutes &
      activity$Calories >= mean_calories, 
      "Highly Active",
  ifelse(
      activity$VeryActiveMinutes >= mean_very_active_minutes |
      activity$Calories >= mean_calories, 
      "Moderately Active",
      "Sedentary"))

table(activity$activity_level)
```

```{r}
# Create a bar plot to visualize each user segment distribution
ggplot(activity, aes(x = activity_level)) +
  geom_bar(fill = "skyblue") +
  labs(title = "Distribution of Activity Levels", x = "Activity Level", y = "Count of Observations")
```

```{r}
# Pie chart of the user segment distribution
ggplot(activity, aes(x="", fill=activity_level))+
  geom_bar(width = 1, color = "white")+
  coord_polar("y") + theme_void()+         scale_fill_brewer(palette = "Set4")
```

By segmenting participants based on the activity metrics, I have created categories with different levels of user engagement with Bellabeat product. The above plots suggest that most of the users tend to spend most of their time not doing much activities or only doing normal daily activities like walking.

I further analyzed when people are most active during the week. The process involves getting the day of week from the ActivityDate and create a bar plot to visualize the frequency by the day of week.

```{r}
Sys.setlocale("LC_TIME", "C")

# Get day of week from the ActivityDate
activity$day_of_week <- weekdays(as.Date(activity$ActivityDate))

activity$day_of_week <- factor(activity$day_of_week,
  levels = c("Sunday", "Monday", 
  "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

```{r}
# Create a bar plot to visualize frequency by day of week
ggplot(activity, aes(x = day_of_week)) +
  geom_bar(fill = "lightpink") +
  labs(title = "Frequency by Day of Week", x = "Day of Week", y = "Frequency")
```

The above bar plot suggests that users are most active on Tuesdays, Wednesdays and Thursdays respectively compared to other days of the week. 

# Act

Based on the analysis and the insights I deduced from the analysis, here are my top three recommendations:

1. Since most of the Bellabeat product users are using the product on a daily basis with sedentary activity, only involving light activity like lying down, sitting with very low energy expenditure, the marketing team should focus on making users understand the importance of exercising for their health and general well-being.

2. The product can highlight features like tracking users' progress by setting activity goals per day and reminders to stay active. It can have a reminder feature that can help users complete their activity each day.

3. The product can also send motivational notifications to encourage users to stay active and improve their fitness habits specifically on week days other than Tuesdays, Wednesdays and Thursdays.
